Ecosystem size and complexity dictate riverine biodiversity

### 2021-03-08 manuscript_science.Rmd
edits: manuscript intro

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR) [@lomolino2000]. In 1921, Arrhenius [@arrhenius1921] formulated the SAR as a power-law $S = cA^z$, an equation currently known as the Arrhenius species-area relationship ($S$ is the number of species observed in a given geographic area $A$, $c$ the constant, and $z$ the scaling exponent). Since then, the spatial scaling of species richness has been observed in many taxonomic groups [@triantis2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity [@kallimanis2008], larger metacommunity size [@hubbell2001], and/or enhanced colonization dynamics [@losos2009]. Importantly, the SAR provides the foundation for global conservation efforts [@myers2000; @desmet2004; @neigel2003]. For example, conservation ecologists have used SAR estimates to design marine and terrestrial protected areas [@desmet2004; @neigel2003], which currently encompass more than 30 million km^2^ globally [@deguignet2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity [@turner2015; @rodríguez-iturbe2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few [@rodríguez-iturbe2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike [@rodríguez-iturbe2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms [@grant2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure [@fagan2002; @grant2010; @brown2010; @henriquessilvarenato], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems [@rinaldo2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches [@carrara2012]. However, most empirical research has explored the consequences of branching complexity for local community structure [@ers2019; @carrara2012] or has relied solely on theoretical arguments with limited replications of branching architecture [@holt2018]. The lack of a metacommunity-level framework integrating the influences of ecosystem size and complexity hinders an objective assessment of their roles in controlling biodiversity patterns. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services [@dudgeon2006].

Here, we show that ecosystem size and complexity play comparable roles in controlling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brown2010; @peterson2013; @moore2015; @terui2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodríguez-iturbe2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@terui2018; @yeakel2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grant2007; @moore2015; @terui2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbell2001; @iwasaki2012]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

>>>>

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR) [@lomolinoEcologyMostGeneral2000]. In 1921, Arrhenius @arrheniusSpeciesArea1921 formulated the SAR as a power-law $S = cA^z$, an equation currently known as the Arrhenius species-area relationship ($S$ is the number of species observed in a given geographic area $A$, $c$ the constant, and $z$ the scaling exponent). Since then, the spatial scaling of species richness has been observed in many taxonomic groups [@triantisIslandSpeciesareaRelationship2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity [@kallimanisHowDoesHabitat2008], larger metacommunity size [@hubbellUnifiedNeutralTheory2001], and/or enhanced colonization dynamics [@lososTheoryIslandBiogeography2009]. Importantly, the SAR provides the foundation for global conservation efforts [@neigelSpeciesareaRelationshipsMarine2003; @desmetUsingSpeciesareaRelationship2004; @myersBiodiversityHotspotsConservation2000]. For example, conservation ecologists have used SAR estimates to design marine and terrestrial protected areas [@neigelSpeciesareaRelationshipsMarine2003; @desmetUsingSpeciesareaRelationship2004], which currently encompass more than 30 million km^2^ globally [@deguignet2014UnitedNations2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity [@turnerLandscapeEcologyTheory2015; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few [@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike [@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms [@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure [@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems [@rinaldoRiverNetworksEcological2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches [@carraraDendriticConnectivityControls2012]. However, most empirical research has explored the consequences of branching complexity for local community structure [@erosLandscapeEcologyRivers2019] or has relied solely on theoretical arguments with limited replications of branching architecture [@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how ecosystem complexity controls biodiversity patterns at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services [@reidEmergingThreatsPersistent2019].

Here, we show that ecosystem size and complexity play comparable roles in controlling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.
